---
title: "FinalProject7.Rmd"
author: "Dana Thacker, Nikole Slinger, Andrew Fahmy, Julia Lundstrum, Zack Larson"
date: "4/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Julia - 2
Andrew - 8
Dana - 7, 3
Nikole - 6, 1
Zack - 4, 5

```{r}
library(ggplot2)
library(tidyverse)
library(readxl)
raw_data <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='raw_data')
VAERSdata <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='2021VAERSDATA')
VAERSsymptoms <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='2021VAERSSymptoms')
VAERSVax <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='2021VAERSVax')
```

**1. What's the most common symptom to experience for each manufacturer?**
```{r}

```

**2. Compare Covid case cases per state in contrast to the number of vaccines administered by state**
```{r}

```

**3. Compare a patient's medical history with their symptoms from the vaccine they received**
```{r}

```

**4. Which states have the highest and lowest number of Covid cases and fatality rate?**
```{r}

```

**5. What patterns are shown when looking at a patient's death (post vaccine) and the reasons why they might have died?**
```{r}

```

**6. What age group has gotten the most vaccines?**
```{r}

```

**7. What is the most popular vaccine manufacturer per state?**
```{r}

```

**8. Is there a trend on the number of vaccines administered per day? By state and nation wide?**
```{r}

```
