---
title: "FinalProject7.Rmd"
author: "Dana Thacker, Nikole Slinger, Andrew Fahmy, Julia Lundstrum, Zack Larson"
date: "4/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1 - Nikole
2 - Julia
3 - Dana
4 - Zack
5 - Zack
6 - Nikole
7 - Dana
8 - Andrew

```{r, warning = FALSE, message = FALSE}
library(ggplot2)
library(tidyverse)
library(readxl)
library(ggthemes)

raw_data <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='raw_data')
VAERSdata <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='2021VAERSDATA')
VAERSsymptoms <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='2021VAERSSymptoms')
VAERSVax <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='2021VAERSVax')
cdcdoses <- readxl::read_xlsx("2021VaccineCovidData.xlsx", sheet='2021AllVaccines')
state <- map_data("state")
abbrev <- read.csv("csvData.csv", stringsAsFactors = FALSE)
```

**1. What's the most common symptom to experience for each manufacturer?**
```{r}
# joining vax and symptom data, only need id, vaccine manu, and symptoms
symptoms <- VAERSVax %>% full_join(VAERSsymptoms, by = 'VAERS_ID') %>%
  filter(VAX_TYPE == 'COVID19') %>%
  select(VAERS_ID, VAX_MANU, SYMPTOM1, SYMPTOM2, SYMPTOM3, SYMPTOM4, SYMPTOM5)

```

**2. Compare Covid case cases per state in contrast to the number of vaccines administered by state**
```{r}
cases <- raw_data %>%
  inner_join(cdcdoses, by = c("Location" = "State")) %>%
  mutate("Cases per 100,000" = `Number of COVID-19 Cases`/`Population`*100000) %>%
  mutate("Doses per 100,000" = `Doses`/`Population`*100000)
casesmap <- cases %>% 
  mutate("State Name" = tolower(`Location`)) %>%
  right_join(state, by = c("State Name" = "region"))
ggplot(casesmap, aes(x=long, y = lat, fill = `Cases per 100,000`)) + geom_polygon(aes(group=group)) + theme_map() + ggtitle("COVID-19 Cases per 100,000 People by State")
ggplot(casesmap, aes(x=long, y = lat, fill = `Doses per 100,000`)) + geom_polygon(aes(group=group)) + theme_map() + ggtitle("Vaccine Doses per 100,000 People by State")
```

**3. Compare a patient's medical history with their symptoms from the vaccine they received.**
```{r}
#Join VAERSsymptoms and VAERSdata, only need ID, all 5 symptoms, and patient history
#Then cleaning the history column to create uniformity
medHist <- VAERSsymptoms %>% full_join(VAERSdata) %>%
  select(VAERS_ID, SYMPTOM1, SYMPTOM2, SYMPTOM3, SYMPTOM4, SYMPTOM5, HISTORY) %>%
  separate(HISTORY, c('history1', 'history2', 'history3', 'history4', 'history5', 'history6', 'history7', 'history8', 
                      'history9', 'history10'), sep = ', ') %>%
  mutate(history1 = tolower(history1), 
         history2 = tolower(history2),
         history3 = tolower(history3),
         history4 = tolower(history4),
         history5 = tolower(history5),
         history6 = tolower(history6),
         history7 = tolower(history7),
         history8 = tolower(history8),
         history9 = tolower(history9),
         history10 = tolower(history10))

#Summarising medical history to find how often each history diagnosis appears in the dataset
medHist2 <- medHist %>%
  pivot_longer(history1:history10, names_to='histName', values_to='conditionHist') %>%
  group_by(conditionHist) %>%
  summarise(count = n())

#Creating a negation of the %in% to use in analysis
`%notin%` <- Negate(`%in%`)

#Finding the top 10, not including 5 null or 'none', medical history conditions
topHist <- top_n(medHist2, 18) %>%
  filter(conditionHist %notin% c('none', 
  'comments: list of non-encoded patient relevant history: patient other relevant history 1: none', 'n/a', 'no',
  'unknown', 'medical history/concurrent conditions: no adverse event (no reported medical history)'))
topHist <- na.omit(topHist)

topHist <- topHist %>%
  mutate(conditionHist = if_else(conditionHist == 'htn', 'hypertension', conditionHist))

topHist[8,2] <- topHist[8,2] + topHist[10,2]

topHist <- topHist[-10, ]

#Filtering the medical history dataset such that I only have the top 10 medical conditions and the covid vaccination symptoms of those conditions
medHist3 <- medHist %>%
  filter(history1 %in% c('anxiety', 'asthma', 'copd', 'depression', 'diabetes', 'gerd', 'high blood pressure',
                         'hypertension', 'htn', 'hyperlipidemia', 'hypothyroidism')) %>%
  select(-history2, -history3, -history4, -history5, -history6, -history7, -history8, -history9, -history10) %>%
  rename('MedHist' = 'history1')

medHist3 <- medHist3 %>%
  mutate(MedHist = if_else(MedHist == 'htn', 'hypertension', MedHist))

#Learning about what the top symptoms are for each medical history
symptom1 <- medHist3 %>%
  group_by(MedHist, SYMPTOM1) %>%
  summarise(symptomCount = n())

symptom2 <- medHist3 %>%
  group_by(MedHist, SYMPTOM2) %>%
  summarise(symptomCount = n())

symptom3 <- medHist3 %>%
  group_by(MedHist, SYMPTOM3) %>%
  summarise(symptomCount = n())

symptom4 <- medHist3 %>%
  group_by(MedHist, SYMPTOM4) %>%
  summarise(symptomCount = n())

symptom5 <- medHist3 %>%
  group_by(MedHist, SYMPTOM5) %>%
  summarise(symptomCount = n())
```

**4. Which states have the highest and lowest number of Covid cases and fatality rate?**
```{r}
raw_data2 <- raw_data
names(raw_data2)<-str_replace_all(names(raw_data2), c(" " = "."))##Column names were not ideal with numbers,commas,dashes
names(raw_data2)<-str_replace_all(names(raw_data2), c("," = "_"))
names(raw_data2)<-str_replace_all(names(raw_data2), c("-19" = ""))
highestCovidCasesStates <- raw_data2
highestCovidCasesStates$COVID.Cases.per.1_000_000.Population <- as.numeric(highestCovidCasesStates$COVID.Cases.per.1_000_000.Population)
highestCovidCasesStates <- highestCovidCasesStates[order(highestCovidCasesStates$COVID.Cases.per.1_000_000.Population,decreasing = TRUE),]
highestCovidCasesStates <- highestCovidCasesStates %>% filter(COVID.Cases.per.1_000_000.Population != 'N/A') %>%
  select(Location,COVID.Cases.per.1_000_000.Population)
highestCovidCasesStates

lowestCovidCasesStates <- raw_data2
lowestCovidCasesStates$COVID.Cases.per.1_000_000.Population <- as.numeric(lowestCovidCasesStates$COVID.Cases.per.1_000_000.Population)
lowestCovidCasesStates <- lowestCovidCasesStates[order(lowestCovidCasesStates$COVID.Cases.per.1_000_000.Population),]
lowestCovidCasesStates <- lowestCovidCasesStates %>% filter(COVID.Cases.per.1_000_000.Population != 'N/A') %>%
  select(Location,COVID.Cases.per.1_000_000.Population)
lowestCovidCasesStates

highestCovidDeathsStates <- raw_data2
highestCovidDeathsStates$COVID.Deaths.per.1_000_000.Population <- as.numeric(highestCovidDeathsStates$COVID.Deaths.per.1_000_000.Population)
highestCovidDeathsStates <- highestCovidDeathsStates[order(highestCovidDeathsStates$COVID.Deaths.per.1_000_000.Population,decreasing = TRUE),]
highestCovidDeathsStates <- highestCovidDeathsStates %>% filter(COVID.Deaths.per.1_000_000.Population != 'N/A') %>%
  select(Location,COVID.Deaths.per.1_000_000.Population)
highestCovidDeathsStates

lowestCovidDeathsStates <- raw_data2
lowestCovidDeathsStates$COVID.Deaths.per.1_000_000.Population <- as.numeric(lowestCovidDeathsStates$COVID.Deaths.per.1_000_000.Population)
lowestCovidDeathsStates <- lowestCovidDeathsStates[order(lowestCovidDeathsStates$COVID.Deaths.per.1_000_000.Population),]
lowestCovidDeathsStates <- lowestCovidDeathsStates %>% filter(COVID.Deaths.per.1_000_000.Population != 'N/A') %>%
  select(Location,COVID.Deaths.per.1_000_000.Population)
lowestCovidDeathsStates 
```

**5. What patterns are shown when looking at a patient's death (post vaccine) and the reasons why they might have died?**
```{r}
#Age
deathAfterVax <- VAERSdata %>% filter(DIED=='Y')
deathAfterVaxAge <- deathAfterVax %>% count(AGE_YRS)
deathAfterVaxAge
ggplot(deathAfterVaxAge,aes(x=AGE_YRS,y=n)) + geom_col()

#Gender
deathAfterVaxSex <- deathAfterVax %>% count(SEX)
deathAfterVaxSex
ggplot(deathAfterVaxSex,aes(x=SEX,y=n)) + geom_col()

#Common symptoms after vax in deaths
deathSymptoms <- deathAfterVax %>% left_join(VAERSsymptoms, by = 'VAERS_ID') %>% group_by(VAERS_ID) %>%
  select(VAERS_ID, AGE_YRS, SEX, SYMPTOM1, SYMPTOM2, SYMPTOM3, SYMPTOM4, SYMPTOM5)

tail(names(sort(table(deathSymptoms$SYMPTOM1))), 5)
tail(names(sort(table(deathSymptoms$SYMPTOM2))), 5)
tail(names(sort(table(deathSymptoms$SYMPTOM3))), 5)
tail(names(sort(table(deathSymptoms$SYMPTOM4))), 5)
tail(names(sort(table(deathSymptoms$SYMPTOM5))), 5)
```

**6. What age group has gotten the most vaccines?**
```{r}
# joining general patient data and vax data
age <- VAERSdata %>% full_join(VAERSVax, by = 'VAERS_ID') %>%
  select(VAERS_ID, VAX_MANU, AGE_YRS)
# adding 10 year age group column to sort ages
age <- age %>%
  mutate(AGE_GROUP = ifelse(AGE_YRS %in% 0:18, '0 to 18',
                     ifelse(AGE_YRS %in% 19:29, '19 to 29',
                     ifelse(AGE_YRS %in% 30:49, '30 to 39',
                     ifelse(AGE_YRS %in% 40:49, '40 to 49',
                     ifelse(AGE_YRS %in% 50:59, '50 to 59',
                     ifelse(AGE_YRS %in% 60:69, '60 to 69', 
                     ifelse(AGE_YRS %in% 70:79, '70 to 79', 
                     ifelse(AGE_YRS %in% 80:89, '80 to 89', 'Over 90')))))))))
# plotting visual representation
ggplot(age, aes(x = AGE_GROUP)) + geom_bar(fill = 'steelblue') +
  ggtitle('Number of Vaccinations per Age Groups') +
  xlab('Age Group (years)') + ylab('Count')
# numerical stats
table(age$AGE_GROUP)
```

**7. What is the most popular vaccine manufacturer per state?**
```{r}
# Joining vaccine dataset with 'data' dataset, only need state, ID number, and vaccine manufacturer
#Summarizing such that the number of each type of vaccine will be counted in each state
perState <- VAERSVax %>% full_join(VAERSdata) %>%
  select(VAERS_ID, STATE, VAX_MANU) %>%
  group_by(STATE, VAX_MANU) %>%
  summarise(popular = n())

#Selecting the top vaccine distributor for each state
topPerState <- top_n(perState, 1) %>%
  select(-popular) 
topPerState <- na.omit(topPerState) %>%
  filter(STATE %notin% c('AS', 'Ca', 'FM', 'GU', 'MH', 'MP', 'PR', 'VI', 'XB'))
head(topPerState, 10)

#Visualizing what manufacturers are the most distributed in different states
ggplot(topPerState, aes(x = VAX_MANU)) + geom_bar() + xlab("Top Vaccine Manufacturers") + ylab("Number of States") + ggtitle("Number of States with Each Vaccine as the Most Distributed Vaccine")

#Creating a Map visualization of which states have each vaccine as the top manufacturer based on number administered
abbrev$State <- toupper(abbrev$State)

state1 <- state %>%
  select(long, lat, region, group)

state1$region <- toupper(state1$region)

stAbbrev <- state1 %>% right_join(abbrev, by = c('region' = 'State'))
stAbbrev <- stAbbrev %>%
  select(-Abbrev)

stAbbrevManu <- stAbbrev %>% right_join(topPerState, by = c('Code' = 'STATE'))

stAbbrevManu <- stAbbrevManu %>%
  mutate(VAX_MANU = as.factor(VAX_MANU))

ggplot(stAbbrevManu, aes(x = long, y = lat)) + geom_polygon(aes(group = group, fill = VAX_MANU)) + xlab("Longitude") + ylab("Latitude") + ggtitle("United States Top Vaccine Manufacturers by Number of Administered Doses per State")
```

The most widely administered vaccine in any given state from the beginning of vaccine administering through March 19, 2021 is either Moderna or Pfizer/Biontech. The most popular vaccine in Iowa as of March 19 was Moderna. Moderna is slightly more common as the most popular vaccine as it was the most popular in 27 states and Pfizer/Biontech was the most popular in 23. 

**8. Is there a trend on the number of vaccines administered per day? By state and nation wide?**
```{r}

```
